Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-03-23 doc/paper/paper.Rnw
Rewrote results section

\documentclass[12pt]{article}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{marginnote}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{multirow}
\usepackage{cleveref}
\usepackage{array}

@@

\newcommand{\set}[1]{\left\lbrace#1\right\rbrace}
\newcommand{\sett}[1]{\{#1\}}
\renewcommand{\emptyset}{\varnothing}
\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\star}[1]{#1^*}
\renewcommand{\d}{\mathrm{d}\,}

@@

\newacronym{BA}{BA}{Barab\'asi-Albert}
\newacronym{SI}{SI}{susceptible-infected}
\newacronym{kSVM}{kSVM}{kernel support vector machine}
\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
\newacronym{SMC}{SMC}{sequential Monte-Carlo}
\newacronym{ABC}{ABC}{approximate Bayesian computation}
\newacronym{MSM}{MSM}{men who have sex with men}
\newacronym{HPD}{HPD}{highest posterior density}
\newacronym{GTR}{GTR}{general time-reversible}
\newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
\newacronym{MAP}{MAP}{maximum \textit{a posteriori}}
\newacronym{IDU}{IDU}{injection drug user}

@@

When an infectious disease spreads through a population, transmissions are
generally more likely to occur between certain pairs of individuals. Such pairs
must have a particular mode of contact with one another, which varies with the
mode of transmission of the disease. For airborne pathogens, physical proximity
may be sufficient, but for sexually transmitted diseases, either sexual or
blood-to-blood contact is required. The set of links between individuals along
which transmission can occur is called the contact
network~\autocite{klovdahl1985social, morris1993epidemiology}. The structure of
the contact network underlying an epidemic can profoundly impact the speed and
pattern of the epidemic's expansion. Network structure can influence the
prevalence trajectory~\autocite{o2010contact} and epidemic
threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
of quantities such as effective population
size~\autocite{goodreau2006assessing}. From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{wang2015targeting}. True contact networks are a challenging
type of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical}.

@@

Survey-based studies of sexual networks~\autocite{liljeros2001web,
schneeberger2004scale} have found that sexual contact networks are best
described by a preferential attachment model \autocite[although there has been
some disagreement, see][]{jones2003assessment}. Under these models, nodes with
a high number of contacts attract new connections at an elevated rate. Networks
produced by preferential attachment have a power-law degree distribution,
meaning that the number of nodes of degree $k$ is proportional to $k^\gamma$
for some constant $\gamma$. These networks are also referred to as
``scale-free''. The first contact network model incorporating preferential
attachment was introduced by \textcite{barabasi1999emergence}, and is now
referred to as the \gls{BA} model. Under this model, networks are formed by
iteratively adding nodes with $m$ new edges each. These new edges are joined to
existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
that nodes of high degree tend to attract more connections (in the original
paper, only $\alpha = 1$ was investigated).

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \textcite{britton2002bayesian} develop a
Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\autocite{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \textcite{brown2011transmission} analysed the degree distribution
of an approximate transmission network, estimated based on genetic similarity
and estimated times of infection, relating 60\% of HIV infected \gls{MSM} in
the United Kingdom. The transmission network is a subgraph of the contact
network which includes only those edges which have already led to a new
infection. The authors found that a Waring distribution, which is produced by a
different preferential attachment model, was a good fit to their estimated
network.

Standard methods of model fitting involve calculation of the likelihood of
observed data under the model. In maximum likelihood estimation, a quantity
proportional to the likelihood is optimized, often through a standard
multi-dimensional numerical optimization procedure. Bayesian methods integrate
prior information by optimizing the posterior probability instead. To avoid
calculation of a normalizing constant, Bayesian inference is often performed
using \gls{MCMC}. Rather than calculating explicit likeilihoods, \gls{MCMC}
uses likelihood \emph{ratios} in which the normalizing constant cancel out.
Unfortunately, it is generally difficult to explicitly calculate the likelihood
of an observed transmission tree under a contact network model, even up to a
normalizing constant. To do so, it would be necessary to integrate over all
possible networks, and also over all possible labellings of the internal nodes
of the transmission tree. While it is not known (to us) whether such
integration is tractable, a simpler alternative is offered by likelihood-free
methods, namely \gls{ABC}. \gls{ABC} leverages the fact that, although
calculating the likelihood may be impossible, generating simulated datasets
according to a model is often straightforward. If our model fits the data well,
the simulated data it produces should be similar to the observed data. More
formally, if $D$ is the observed data, the posterior distribution $f(\theta
\mid D)$ on model parameters $\theta$ is replaced as the target of statistical
inference by $f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a
distance function, $\hat{D}$ is a simulated dataset according to $\theta$, and
$\varepsilon$ is a small tolerance~\autocite{sunnaker2013approximate}. In the
specific case when $\rho$ is a kernel function, the approach is known as
kernel-\gls{ABC}~\autocite{nakagome2013kernel, poon2015phylodynamic}.

@@

We chose to study the \gls{BA} network model~\autocite{barabasi1999emergence}.
In addition to $m$ and $\alpha$, we investigated the parameters $N$, which
denotes the total number of nodes in the network, and $I$, which is the number
of infected nodes at which to stop the simulation and sample the transmission
tree. Nodes in our networks followed simple \gls{SI} dynamics, meaning that
they became infected at a rate proportional to their numbers of infected
neighbours, and never recovered. For all analyses, the transmission trees'
branch lengths were scaled by dividing by their mean. We used the 
\software{igraph} library's implementation of the \gls{BA}
model~\autocite{csardi2006igraph} to generate the graphs. The analyses were run
on Westgrid (\url{https://www.westgrid.ca/}) and a local computer cluster.

@@

Three similar experiments were performed for the other \gls{BA} model
parameters (one experiment per parameter). $m$ was varied between 2, 3, and 4;
$I$ between 500, 1000, and 2000; and $N$ between 3000, 5000, and 8000. The
parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
$m$ = 2, and $\alpha$ = 1. Thus, we performed a total of four kSVM
cross-validations, one for each of the \gls{BA} model parameters $\alpha$, $I$,
$m$, and $N$.

We repeated these four cross-validations with different values of $\lambda$
(0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well
as on trees with differing numbers of tips (100, 500, and 1000) and in
epidemics of differing size $I$ (500, 1000, and 2000). When evaluating the
classifier for $I$, we did not consider trees with 1000 tips, because one of
the tested $I$ values was 500, and the number of tips cannot be larger than
$I$.

@@

We simulated three transmission trees, each with 500 tips, under every element
of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1,
1.5}. This produced a total of 24 parameter combinations $\times$ three trees
per combination = 72 trees total. The adaptive \gls{ABC} algorithm was applied
to each tree with these priors: $m \sim$ Uniform(1, 5), $\alpha \sim$
Uniform(0, 2), and $(N, I)$ jointly uniform on the triangular region \{$500
\leq N \leq 15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Following
\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, all proposals
were Gaussian, with variance equal to twice the empirical variance of the
particles. The algorithm was run with 1000 particles, 5 simulated datasets per
particle, and the ``quality'' parameter controlling the decay rate of the
tolerance $\varepsilon$ set to 0.95. We used the same stopping criterion as
\citeauthor{del2012adaptive}, namely when the \gls{MCMC} acceptance rate
dropped below 1.5\%. Point estimates for the parameters were obtained by taking
the highest point of an estimated kernel density on the final set of particles,
using the \software{density} function with the default parameters in
\software{R}. \gls{HPD} intervals were calculated with the
\software{HPDinterval} function from the \software{R} package
\software{coda}~\autocite{plummer2006coda}.

@@

We downloaded all sequences associated with each study from GenBank. For the
\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the
HXB2 reference sequence (Genbank accession number HIVHXB2CG) and the
hypervariable regions were clipped out with BioPython version
1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
\software{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments
were manually inspected with \software{Seaview} version 4.4.2
\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
alignments by approximate maximum likelihood using \software{FastTree2} version
2.1.7 with the \gls{GTR} model. Transmission trees were estimated by rooting
and time-scaling the phylogenies by root-to-tip regression, using a modified
version of Path-O-Gen (distributed as part of
BEAST~\autocite{drummond2007beast}) as described
previously~\autocite{poon2015phylodynamic}.

@@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested~\cref{fig:rsquared}. Classifiers based on two other tree statistics, the
\gls{nLTT} and Sackin's index, generally exhibited worse performance than the
tree kernel, although the magnitude of the disparity varied between the
parameters. \Cref{fig:rsquared} shows the cross-validation accuracy with
$\lambda$ = 0.3, $\sigma$ = 4, $I$ = 1000, and trees of size 500. The results
were largely robust to variations in $\lambda$ and $\sigma$, although accuracy
varied between different epidemic and sampling scenarios
(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).

The kSVM classifier for $\alpha$ had an average $R^2$ of
    \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]} 
(\cref{fig:alphacrossv}). The $R^2$ of the for the SVM classifer for $\alpha$
using the \gls{nLTT} varied between
\Sexpr{nltt.alpha[,round(min(rsquared), 2)]}
and
  \Sexpr{nltt.alpha[,round(max(rsquared), 2)]}.
The $R^2$ of the linear regression against Sackin's index varied between
  \Sexpr{sackin.alpha[,round(min(rsquared), 2)]}
and
  \Sexpr{sackin.alpha[,round(max(rsquared), 2)]}.
The $I$ parameter was also classified accurately by the tree kernel
(\cref{fig:Icrossv}), with an average
$R^2$ of 
  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]} 
for 500-tip trees or 
  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]} 
for 100-tip trees. The \gls{nLTT} performed almost as well on this parameter as
the tree kernel, with $R^2$ values of
  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}
for 500-tip trees or
  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}
for 100-tip trees. Sackin's index was an extremely poor classifier for this
parameter, with both $R^2$ values below
  \Sexpr{sackin.I[,round(max(rsquared) + 0.01, 2)]}.

The $m$ parameter was much harder to classify (\cref{fig:mcrossv}), with $R^2$
for the kSVM varying between 
  \Sexpr{kernel.m[,round(min(rsquared), 3)]}
and
  \Sexpr{kernel.m[,round(max(rsquared), 2)]}
depending on $I$ and the tree size. The \gls{nLTT} was also universally poor,
with $R^2$ below
  \Sexpr{nltt.m[,round(max(rsquared) + 0.01, 2)]}
for all scenarios. Sackin's index fared slightly better than the \gls{nLTT},
especially on trees with a higher number of tips, with $R^2$ between
  \Sexpr{sackin.m[,round(min(rsquared), 2)]}
and
  \Sexpr{sackin.m[,round(max(rsquared), 2)]}.
Finally, the accuracy of the kSVM classifier for $N$ varied widely
(\cref{fig:Ncrossv}), from 
  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
to
  \Sexpr{kernel.N[,round(max(rsquared), 2)]}.
The \gls{nLTT} performed well in classifying $N$ when the epidemic was large,
improving upon the tree kernel in the scenarios with $I$ = 2000 and 100- or
500-tip transmission trees ($R^2$ = 
  \Sexpr{nltt.N[which(I == 2000 & ntip == 100), round(mean(rsquared), 2)]}
and
  \Sexpr{nltt.N[which(I == 2000 & ntip == 500), round(mean(rsquared), 2)]}),
although it performed poorly when $I$ was 500 or 1000 (average $R^2$ =
  \Sexpr{nltt.N[which(I < 2000), round(mean(rsquared), 2)]}).
Sackin's index could not classify $N$, with all $R^2$ below
  \Sexpr{sackin.N[,round(max(rsquared) + 0.01, 2)]}.
Based on inspection of the cross-validation results, we chose to use the
meta-parameters $\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not
to use Sackin's index or the \gls{nLTT}.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
  \caption{
      Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
      using \gls{nLTT} (centre), and linear regression using Sackin's index
      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
      simulated transmission trees over networks with three different values of
      the parameter being tested. Vertical lines are empirical 95\% confidence
      intervals based on 1000 2-fold cross-validations.
  }
  \label{fig:rsquared}
\end{figure}

@@

    m.tbl <- d[,prop.table(table(m_error))]
    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
                                   d[true_alpha != 0, alpha_error])
    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
                                d[true_alpha >= 1, I_error])
    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
                            d[true_I == 2000, I_error])
    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])

    options(scipen=-1, digits=2)
@

>>>>

\documentclass[12pt]{article}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{marginnote}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{multirow}
\usepackage{cleveref}
\usepackage{array}
\usepackage{float}

\newfloat{data}{hb}{lod}
\floatname{data}{Data}
\crefname{data}{data}{data}

@@

\newcommand{\set}[1]{\left\lbrace#1\right\rbrace}
\newcommand{\sett}[1]{\{#1\}}
\newcommand{\ie}{\textit{i.e.}}
\renewcommand{\emptyset}{\varnothing}
\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\star}[1]{#1^*}
\renewcommand{\d}{\mathrm{d}\,}

@@

\newacronym{BA}{BA}{Barab\'asi-Albert}
\newacronym{SI}{SI}{susceptible-infected}
\newacronym{kSVM}{kSVM}{kernel support vector machine}
\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
\newacronym{SMC}{SMC}{sequential Monte-Carlo}
\newacronym{ABC}{ABC}{approximate Bayesian computation}
\newacronym{MSM}{MSM}{men who have sex with men}
\newacronym{HPD}{HPD}{highest posterior density}
\newacronym{GTR}{GTR}{general time-reversible}
\newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
\newacronym{MAP}{MAP}{maximum \textit{a posteriori}}
\newacronym{IDU}{IDU}{injection drug user}
\newacronym{HIV}{HIV}{human immuno-deficiency virus}

@@

When an infectious disease spreads through a population, transmissions are
generally more likely to occur between certain pairs of individuals. Such pairs
must have a particular mode of contact with one another, which varies with the
mode of transmission of the disease. For airborne pathogens, physical proximity
may be sufficient, while for sexually transmitted diseases, sexual or in some
cases blood-to-blood contact is required. The population together with the set
of links between individuals along which transmission can occur is called the
contact network~\autocite{klovdahl1985social, morris1993epidemiology}. The
structure of the contact network underlying an epidemic can profoundly impact
the speed and pattern of the epidemic's expansion. Network structure can
influence the prevalence trajectory~\autocite{o2010contact} and epidemic
threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
of quantities such as effective population
size~\autocite{goodreau2006assessing}. From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{wang2015targeting}. True contact networks are a challenging
type of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical}.

@@

Survey-based studies of sexual networks~\autocite{liljeros2001web,
schneeberger2004scale, colgate1989risk} have found that these networks have a
degree distribution which follows a power law \autocite[although there has been
some disagreement, see][]{handcock2004likelihood}. That is, number of nodes of
degree $k$ is proportional to $k^-\gamma$ for some constant $\gamma$. These
networks are also referred to as ``scale-free''. One process by which
scale-free networks can be generated is preferential attachment, where nodes
with a high number of contacts attract new connections at an elevated rate. The
first contact network model incorporating preferential attachment was
introduced by \textcite{barabasi1999emergence}, and is now referred to as the
\gls{BA} model. Under this model, networks are formed by iteratively adding
nodes with $m$ new edges each. In the most commonly studied formulation, these
new edges are joined to existing nodes of degree $k$ with probability
proportional to $k$, so that nodes of high degree tend to attract more
connections. \citeauthor{barabasi1999emergence} suggested an extension where
the probability of attaching to a node of degree $k$ is $k^\alpha$ for some
non-negative constant $\alpha$, and we use this extension here.

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \textcite{britton2002bayesian} develop a
Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\autocite{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \textcite{volz2007susceptible, volz2008sir} developed differential
equations describing the spread of a \gls{SI} epidemic on static and dynamic
contact networks with several degree distributions, which could in principle be
used for inference if observed incidence trajectories were available.
\textcite{brown2011transmission} analysed the degree distribution of an
approximate transmission network, estimated based on genetic similarity and
estimated times of infection, relating 60\% of \gls{HIV}-infected \gls{MSM} in
the United Kingdom. The transmission network is a subgraph of the contact
network which includes only those edges which have already led to a new
infection. The authors found that a Waring distribution, which is produced by a
more sophisticated preferential attachment model, was a good fit to their
estimated network.

Standard methods of model fitting involve calculation of the likelihood of
observed data under the model. In maximum likelihood estimation, a quantity
proportional to the likelihood is optimized, often through a standard
multi-dimensional numerical optimization procedure. Bayesian methods integrate
prior information by optimizing the posterior probability instead. To avoid
calculation of a normalizing constant, Bayesian inference is often performed
using \gls{MCMC}, which uses likelihood \emph{ratios} in which the normalizing
constants cancel out. Unfortunately, it is generally difficult to explicitly
calculate the likelihood of an observed transmission tree under a contact
network model, even up to a normalizing constant. To do so, it would be
necessary to integrate over all possible networks, and also over all possible
labellings of the internal nodes of the transmission tree. While it is not
known (to us) whether such integration is tractable, a simpler alternative is
offered by likelihood-free methods, namely \gls{ABC}. \gls{ABC} leverages the
fact that, although calculating the likelihood may be impractical, generating
simulated datasets according to a model is often straightforward. If our model
fits the data well, the simulated data it produces should be similar to the
observed data. More formally, if $D$ is the observed data, the posterior
distribution $f(\theta \mid D)$ on model parameters $\theta$ is replaced as the
target of statistical inference by $f(\theta \mid \rho(\hat{D}, D) <
\varepsilon)$, where $\rho$ is a distance function, $\hat{D}$ is a simulated
dataset according to $\theta$, and $\varepsilon$ is a small
tolerance~\autocite{sunnaker2013approximate}. In the specific case when $\rho$
is a kernel function, the approach is known as
kernel-\gls{ABC}~\autocite{nakagome2013kernel, poon2015phylodynamic}.

@@

We chose to study the \gls{BA} network model~\autocite{barabasi1999emergence}.
In addition to $m$ and $\alpha$, we investigated the parameters $N$, which
denotes the total number of nodes in the network, and $I$, which is the number
of infected nodes at which to stop the simulation and sample the transmission
tree. Nodes in our networks followed simple \gls{SI} dynamics, meaning that
they became infected at a rate proportional to their number of infected
neighbours, and never recovered. For all analyses, the transmission trees'
branch lengths were scaled by dividing by their mean. We used the 
\software{igraph} library's implementation of the \gls{BA}
model~\autocite{csardi2006igraph} to generate the graphs. The analyses were run
on Westgrid (\url{https://www.westgrid.ca/}) and a local computer cluster.

@@

Three similar experiments were performed for the other \gls{BA} model
parameters (one experiment per parameter). $m$ was varied between 2, 3, and 4;
$I$ between 500, 1000, and 2000; and $N$ between 3000, 5000, and 8000. The
parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
$m$ = 2, and $\alpha$ = 1. Thus, we performed a total of four kSVM
cross-validations, one for each of the \gls{BA} model parameters $\alpha$, $I$,
$m$, and $N$. We repeated these four cross-validations with different values of
$\lambda$ (0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$),
as well as on trees with differing numbers of tips (100, 500, and 1000) and in
epidemics of differing size (500, 1000, and 2000). When evaluating the
classifier for $I$, we did not consider trees with 1000 tips, because one of
the tested $I$ values was 500, and the number of tips cannot be larger than
$I$.

@@

We simulated three transmission trees, each with 500 tips, under every element
of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1,
1.5}. This produced a total of 24 parameter combinations $\times$ three trees
per combination = 72 trees total. The adaptive \gls{ABC} algorithm was applied
to each tree with these priors: $m \sim$ Uniform(1, 5), $\alpha \sim$
Uniform(0, 2), and $(N, I)$ jointly uniform on the region \{$500 \leq N \leq
15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Following
\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, all proposals
were Gaussian, with variance equal to twice the empirical variance of the
particles. The algorithm was run with 1000 particles, 5 simulated datasets per
particle, and the ``quality'' parameter controlling the decay rate of the
tolerance $\varepsilon$ set to 0.95. We used the same stopping criterion as
\citeauthor{del2012adaptive}, namely when the \gls{MCMC} acceptance rate
dropped below 1.5\%. Point estimates for the parameters were obtained by taking
the highest point of an estimated kernel density on the final set of particles,
using the \software{density} function with the default parameters in
\software{R}. \Gls{HPD} intervals were calculated with the
\software{HPDinterval} function from the \software{R} package
\software{coda}~\autocite{plummer2006coda}.

@@

We downloaded all sequences associated with each study from GenBank. For the
\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the
HXB2 reference sequence (Genbank accession number HIVHXB2CG) and the
hypervariable regions were clipped out with \software{BioPython} version
1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
\software{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments
were manually inspected with \software{Seaview} version 4.4.2
\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
alignments by approximate maximum likelihood using \software{FastTree2} version
2.1.7 with the \gls{GTR} model. Transmission trees were estimated by rooting
and time-scaling the phylogenies by root-to-tip regression, using a modified
version of Path-O-Gen (distributed as part of
BEAST~\autocite{drummond2007beast}) as described
previously~\autocite{poon2015phylodynamic}.

@@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested (\cref{fig:rsquared}, left). Classifiers based on two other tree
statistics, the \gls{nLTT} and Sackin's index, generally exhibited worse
performance than the tree kernel, although the magnitude of the disparity
varied between the parameters (\cref{fig:rsquared}, centre and right).
\Cref{fig:rsquared} shows the cross-validation accuracy of each of the tree
classifiers. The results were largely robust to variations in the tree kernel
meta-parameters $\lambda$ and $\sigma$, although accuracy varied between
different epidemic and sampling scenarios
(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).

When classifying $\alpha$, the kernel-SVM classifier had an average $R^2$ of 
    \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
compared to 
    \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
for the \gls{nLTT}-based SVM, and
    \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
for the linear regression against Sackin's index. There was little variation
about the mean for different tree and epidemic sizes. No classifier could
accurately identify the $m$ parameter in any epidemic scenario, with average
$R^2$ values of 
  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for \gls{kSVM},
  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the \gls{nLTT}, and
  \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
for Sackin's index. Again, there was little variation in accuracy between
epidemic scenarios, although the accuracy of the \gls{kSVM} was slightly higher
on 1000-tip trees.

The accuracy of classifiers $I$ varied significantly with the number of tips in
the tree. For 100-tip trees, the average $R^2$ values were
  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
  \Sexpr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
for the tree kernel, \gls{nLTT}, and Sackin's index respectively. For 500-tip
trees, the values increased to
  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
  \Sexpr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
Finally, the performance of classifiers for $N$ depended heavily on the
epidemic scenario. The $R^2$ of the \gls{kSVM} classifier ranged from
  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
for the smallest epidemic and smallest sample size, to
  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
for the largest. Likewise, $R^2$ for the \gls{nLTT}-based SVM ranged from 
  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
to
  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
Sackin's index did not accurately classify $N$ in any scenario, with an average
$R^2$ of
  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
and little variation between scenarios.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
  \caption{
      Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
      using \gls{nLTT} (centre), and linear regression using Sackin's index
      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
      simulated transmission trees over networks with three different values of
      the parameter being tested. Vertical lines are empirical 95\% confidence
      intervals based on 1000 two-fold cross-validations.
  }
  \label{fig:rsquared}
\end{figure}

@@

    m.tbl <- d[,prop.table(table(m_error))]
    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
                                   d[true_alpha != 0, alpha_error])
    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
                                d[true_alpha >= 1, I_error])
    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
                            d[true_I == 2000, I_error])
    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
    options(scipen=4)
    stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] > 0.05)
    stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
    stopifnot(I.I.test$p.value < 1e-5)
    stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
    stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
    stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
    stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
@

\Cref{fig:abcpt} shows \gls{MAP} point estimates of the \gls{BA} model
parameters obtained with kernel-\gls{ABC} on simulated data. The estimates
shown correspond only to the simulations where the $m$ parameter was set to 2,
however the results for $m = 3$ and $m = 4$ were similar 
(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% \gls{HPD}
intervals are given in \cref{tab:abchpd}.

The accuracy of the parameter estimates obtained with kernel-\gls{ABC}
paralleled the results from the \gls{kSVM} classifier. Of the four parameters,
$\alpha$ was the most accurately estimated, with point estimates having a
median [IQR] absolute error of 
    \Sexpr{d[,round(median(alpha_error), 2)]} 
    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}].
The errors when the true value of $\alpha$ was zero were significantly greater
than those for the other values 
    (Wilcoxon rank-sum test, $p$ = \Sexpr{round(zero.alpha.test$p.value, 4)}),
but did not vary across the true values of the other parameters (one-way
ANOVA). Estimates for $I$ were also relatively accurate, with point estimate
errors of
    \Sexpr{d[,round(median(I_error))]} 
    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
    \Sexpr{d[,round(quantile(I_error, 0.75))]}].
These errors were significantly higher when the true value of $\alpha$ was
at least 1
    (Wilcoxon rank-sum test, $p$ = \Sexpr{round(alpha.I.test$p.value, 4)})
and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
did not affect the estimates of $I$ (one-way ANOVA).

The $m$ parameter was estimated correctly in only
    \Sexpr{as.integer(m.tbl[1] * 100)} \%
of simulations. Oddly, the error in the estimated $m$ was higher for integer
values of $\alpha$ (\ie~0 and 1) than non-integer values 
    (Wilcoxon rask-sum test, $p$ = \Sexpr{round(alpha.m.test$p.value, 3)}).
The true values of the other parameters did not significantly affect the
estimates of $m$ (both one-way ANOVA). Finally, the total number of nodes $N$
was consistently over-estimated by about a factor of two
    (error \Sexpr{d[,round(median(N_error))]} 
    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
     \Sexpr{d[,round(quantile(N_error, 0.75))]}]).
No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
